Machine Learning and the Many-body Problem

### 2019-01-17 doc/MachineLearning/article1.tex
update

\pacs{32.80.Rm, 31.15.A-, 42.65.Ky}

@@

\section{Introduction}

%%
%% Automatically generated file from DocOnce source
%% (https://github.com/hplgit/doconce/)
%%
%%

%-------------------- begin preamble ----------------------

\documentclass[%
oneside,                 % oneside: electronic viewing, twoside: printing
final,                   % draft: marks overfull hboxes, figures with paths
10pt]{article}

\listfiles               %  print all files needed to compile this document

\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
\usepackage[table]{xcolor}
\usepackage{bm,ltablex,microtype}

\usepackage[pdftex]{graphicx}

\usepackage{fancyvrb} % packages needed for verbatim environments
\usepackage{minted}
\usemintedstyle{default}

\usepackage[T1]{fontenc}
%\usepackage[latin1]{inputenc}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}

\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern

% Hyperlinks in PDF:
\definecolor{linkcolor}{rgb}{0,0,0.4}
\usepackage{hyperref}
\hypersetup{
    breaklinks=true,
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    citecolor=black,
    filecolor=black,
    %filecolor=blue,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
    }
%\hyperbaseurl{}   % hyperlinks are relative to this root

\setcounter{tocdepth}{2}  % levels in table of contents

% Tricks for having figures close to where they are defined:
% 1. define less restrictive rules for where to put figures
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{0.75}
% floatpagefraction must always be less than topfraction!
% 2. ensure all figures are flushed before next section
\usepackage[section]{placeins}
% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
%\usepackage{float}\restylefloat{figure}

% --- fancyhdr package for fancy headers ---
\usepackage{fancyhdr}
\fancyhf{} % sets both header and footer to nothing
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[LE,RO]{\thepage}
% Ensure copyright on titlepage (article style) and chapter pages (book style)
\fancypagestyle{plain}{
  \fancyhf{}
  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
%  \renewcommand{\footrulewidth}{0mm}
  \renewcommand{\headrulewidth}{0mm}
}
% Ensure copyright on titlepages with \thispagestyle{empty}
\fancypagestyle{empty}{
  \fancyhf{}
  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
  \renewcommand{\footrulewidth}{0mm}
  \renewcommand{\headrulewidth}{0mm}
}

\pagestyle{fancy}

\usepackage[framemethod=TikZ]{mdframed}

% --- begin definitions of admonition environments ---

% --- end of definitions of admonition environments ---

% prevent orhpans and widows
\clubpenalty = 10000
\widowpenalty = 10000

% --- end of standard preamble for documents ---

% insert custom LaTeX commands...

\raggedbottom
\makeindex
\usepackage[totoc]{idxlayout}   % for index in the toc
\usepackage[nottoc]{tocbibind}  % for references/bibliography in the toc

%-------------------- end preamble ----------------------

% matching end for #ifdef PREAMBLE

\newcommand{\exercisesection}[1]{\subsection*{#1}}

% ------------------- main content ----------------------

% ----------------- title -------------------------

\thispagestyle{empty}

\begin{center}
{\LARGE\bf
\begin{spacing}{1.25}
Solving quantum mechanical problems with Machine Learning
\end{spacing}
}
\end{center}

% ----------------- author(s) -------------------------

\begin{center}
{\bf Morten Hjorth-Jensen${}^{1, 2}$} \\ [0mm]
\end{center}

\begin{center}
% List of all institutions:
\centerline{{\small ${}^1$Department of Physics and Astronomy and National Superconducting Cyclotron Laboratory, Michigan State University, USA}}
\centerline{{\small ${}^2$Department of Physics and Center for Computing in Science Education, University of Oslo, Norway}}
\end{center}

% ----------------- end author(s) -------------------------

% --- begin date ---
\begin{center}
Jan 16, 2019 
\end{center}
% --- end date ---

\vspace{1cm}

% !split
\subsection*{What is this talk about?}

% --- begin paragraph admon ---
\paragraph{}
The main aim is to give you a short and pedestrian introduction to  how we can use Machine Learning methods
to solve quantum mechanical many-body problems. And why this could be of interest.

The hope is that after this talk you have gotten the basic ideas to get you started. Peeping into \href{{https://github.com/mhjensenseminars/MachineLearningTalk}}{\nolinkurl{https://github.com/mhjensenseminars/MachineLearningTalk}}, you'll find a Jupyter notebook, slides, codes etc that will allow you to reproduce the simulations discussed here, and perhaps run your own very first calculations.

Try it out and please don't hesitate to swing by if something is unclear.
% --- end paragraph admon ---

% !split
\subsection*{Why?}

How can we avoid the dimensionality curse? Many possibilities
\begin{enumerate}
\item smarter basis functions

\item resummation of specific correlations

\item stochastic sampling of high-lying states (stochastic FCI, CC and SRG/IMSRG)

\item many more
\end{enumerate}

\noindent
Machine Learning and Quantum Computing hold also great promise in tackling the 
ever increasing dimensionalities. Here we will focus on Machine Learning.

% !split
\subsection*{Overview}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
\item Short intro to Machine Learning

\item Variational Monte Carlo (Markov Chain Monte Carlo, $\mathrm{MC}^2$) and two-electron quantum dots, solving quantum mechanical problems in a stochastic way. It  will serve as  our motivation for switching to Machine Learning.

\item From Variational Monte Carlo to Boltzmann Machines and Machine Learning
\end{itemize}

% !split
\subsection*{What are the Machine Learning calculations here based on?}

@@

Thanks to Jane Kim (MSU), Vilde Flugsrud (UiO), Alfred Alocias Mariadason (UiO) for many discussions and interpretations.

% !split
\subsection*{A new world}

% --- begin paragraph admon ---
\paragraph{}
Machine learning (ML) is an extremely rich field, in spite of its young age. The
increases we have seen during the last three decades in computational
capabilities have been followed by developments of methods and
techniques for analyzing and handling large date sets, relying heavily
on statistics, computer science and mathematics.  The field is rather
new and developing rapidly.

Popular software packages written in Python for ML are

\begin{itemize}
\item \href{{http://scikit-learn.org/stable/}}{Scikit-learn},

\item \href{{https://www.tensorflow.org/}}{Tensorflow},

\item \href{{http://pytorch.org/}}{PyTorch} and

\item \href{{https://keras.io/}}{Keras}.
\end{itemize}

\noindent
These are all freely available at their respective GitHub sites. They 
encompass communities of developers in the thousands or more. And the number
of code developers and contributors keeps increasing.
% --- end paragraph admon ---

% !split
\subsection*{Lots of room for creativity}

% --- begin paragraph admon ---
\paragraph{}
Not all the
algorithms and methods can be given a rigorous mathematical
justification, opening up thereby for experimenting
and trial and error and thereby exciting new developments.
% --- end paragraph admon ---

% --- begin paragraph admon ---
\paragraph{}
A solid command of linear algebra, multivariate theory, 
probability theory, statistical data analysis,
understanding errors and Monte Carlo methods is important in order to understand many of the 
various algorithms and methods.
% --- end paragraph admon ---

\textbf{Job market, a personal statement}: \href{{https://www.analyticsindiamag.com/top-countries-hiring-most-number-of-artificial-intelligence-machine-learning-experts/}}{A familiarity with ML is almost becoming a prerequisite for many of the most exciting employment opportunities}. And add quantum computing and there you are!

% !split
\subsection*{Knowledge of Statistical analysis and optimization of data}
Some key elements that enter much of the discussion on ML:
\begin{enumerate}
\item Basic concepts, expectation values, variance, covariance, correlation functions and errors;

\item Simpler models, binomial distribution, the Poisson distribution, simple and multivariate normal distributions;

\item Central elements of Bayesian statistics and modeling;

\item Central elements from linear algebra

\item Gradient methods for data optimization

\item Monte Carlo methods, Markov chains, Metropolis-Hastings algorithm;

\item Estimation of errors using cross-validation, blocking, bootstrapping and jackknife methods;

\item Practical optimization using Singular-value decomposition and least squares for parameterizing data.

\item Principal Component Analysis.
\end{enumerate}

\noindent
% !split 
\subsection*{Some members of the ML family}

% --- begin paragraph admon ---
\paragraph{}
\begin{enumerate}
\item Linear regression and its variants, Logistic regression

\item Decision tree algorithms, from simpler to more complex ones like random forests

\item Bayesian statistics

\item Support vector machines and finally various variants of

\item Artifical neural networks and deep learning

\item Convolutional NN, autoenconders

\item and many more
\end{enumerate}

% !split
\subsection*{What are the basic ingredients?}

% --- begin paragraph admon ---
\paragraph{}
Almost every problem in ML and data science starts with the same ingredients:
\begin{itemize}
\item The dataset $\mathbf{x}$ (could be some observable quantity of the system we are studying)

@@

\noindent
We seek to minimize the function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ by finding the parameter values which minimize $\mathcal{C}$. This leads to  various minimization algorithms.
% --- end paragraph admon ---

% !split
\subsection*{What is Machine Learning?}

% --- begin paragraph admon ---
\paragraph{}
Machine learning is the science of giving computers the ability to
learn without being explicitly programmed.  The idea is that there
exist generic algorithms which can be used to find patterns in a broad
class of data sets without having to write code specifically for each
problem. The algorithm will build its own logic based on the data.

@@

You will notice however that many of the basic ideas discussed do come from Physics!
% --- end paragraph admon ---

% !split
\subsection*{Types of Machine Learning}

% --- begin paragraph admon ---
\paragraph{}
The approaches to machine learning are many, but are often split into two main categories. 
In \emph{supervised learning} we know the answer to a problem,
and let the computer deduce the logic behind it. On the other hand, \emph{unsupervised learning}
is a method for finding patterns and relationship in data sets without any prior knowledge of the system.
Some authours also operate with a third category, namely \emph{reinforcement learning}. This is a paradigm 
of learning inspired by behavioural psychology, where learning is achieved by trial-and-error, 
solely from rewards and punishment.

@@

% !split
\subsection*{References}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
\item An excellent reference, \href{{https://arxiv.org/abs/1803.08823}}{Mehta \emph{et al.}, arXiv:1803.08823 and Physics Reports in press (2018)}

@@

\item See also Metha and Schwab, arXiv.1410.3831, how to link Variational renormalization group theory with deep learning (recommended read)
\end{itemize}

% !split
\subsection*{Another interesting article}

@@

% !split
\subsection*{Courses on Data science and Machine Learning  at UiO}

The \href{{https://www.mn.uio.no/english/research/about/centre-focus/innovation/data-science/studies/}}{link here}  gives an excellent overview of courses on Machine learning at UiO.

\begin{enumerate}
\item \href{{http://www.uio.no/studier/emner/matnat/math/STK2100/index-eng.html}}{STK2100 Machine learning and statistical methods for prediction and classification}.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN3050/index-eng.html}}{IN3050 Introduction to Artificial Intelligence and Machine Learning}. Introductory course in machine learning and AI with an algorithmic approach.

\item \href{{http://www.uio.no/studier/emner/matnat/math/STK-INF3000/index-eng.html}}{STK-INF3000/4000 Selected Topics in Data Science}. The course provides insight into selected contemporary relevant topics within Data Science.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN4080/index.html}}{IN4080 Natural Language Processing}. Probabilistic and machine learning techniques applied to natural language processing.

\item \href{{https://www.uio.no/studier/emner/matnat/math/STK-IN4300/index-eng.html}}{STK-IN4300 – Statistical learning methods in Data Science}. An advanced introduction to statistical and machine learning. For students with a good mathematics and statistics background.

\item \href{{http://www.uio.no/studier/emner/matnat/ifi/INF4490/}}{INF4490 Biologically Inspired Computing}. An introduction to self-adapting methods also called artificial intelligence or machine learning.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN-STK5000/index-eng.html}}{IN-STK5000  Adaptive Methods for Data-Based Decision Making}. Methods for adaptive collection and processing of data based on machine learning techniques.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN5400/}}{IN5400/INF5860 – Machine Learning for Image Analysis}. An introduction to deep learning with particular emphasis on applications within Image analysis, but useful for other application areas too.

\item \href{{https://www.uio.no/studier/emner/matnat/its/TEK5040/}}{TEK5040 – Deep learning}. The course addresses advanced algorithms and architectures for deep learning with neural networks. The course provides an introduction to how deep-learning techniques can be used in the construction of key parts of advanced autonomous systems that exist in physical environments and cyber environments.
\end{enumerate}

\noindent
% !split
\subsection*{Additional courses of interest}

\begin{enumerate}
\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4051/index-eng.html}}{STK4051 Computational Statistics}

\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4021/index-eng.html}}{STK4021 Applied Bayesian Analysis and Numerical Methods}
\end{enumerate}

\noindent
% !split
\subsection*{Decision trees and Regression}
\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
import numpy as np
import matplotlib.pyplot as plt
from sklearn.preprocessing import PolynomialFeatures
from sklearn.linear_model import LinearRegression

steps=250

distance=0
x=0
distance_list=[]
steps_list=[]
while x<steps:
    distance+=np.random.randint(-1,2)
    distance_list.append(distance)
    x+=1
    steps_list.append(x)
plt.plot(steps_list,distance_list, color='green', label="Random Walk Data")

steps_list=np.asarray(steps_list)
distance_list=np.asarray(distance_list)

X=steps_list[:,np.newaxis]

#Polynomial fits

#Degree 2
poly_features=PolynomialFeatures(degree=2, include_bias=False)
X_poly=poly_features.fit_transform(X)

lin_reg=LinearRegression()
poly_fit=lin_reg.fit(X_poly,distance_list)
b=lin_reg.coef_
c=lin_reg.intercept_
print ("2nd degree coefficients:")
print ("zero power: ",c)
print ("first power: ", b[0])
print ("second power: ",b[1])

z = np.arange(0, steps, .01)
z_mod=b[1]*z**2+b[0]*z+c

fit_mod=b[1]*X**2+b[0]*X+c
plt.plot(z, z_mod, color='r', label="2nd Degree Fit")
plt.title("Polynomial Regression")

plt.xlabel("Steps")
plt.ylabel("Distance")

#Degree 10
poly_features10=PolynomialFeatures(degree=10, include_bias=False)
X_poly10=poly_features10.fit_transform(X)

poly_fit10=lin_reg.fit(X_poly10,distance_list)

y_plot=poly_fit10.predict(X_poly10)
plt.plot(X, y_plot, color='black', label="10th Degree Fit")

plt.legend()
plt.show()

#Decision Tree Regression
from sklearn.tree import DecisionTreeRegressor
regr_1=DecisionTreeRegressor(max_depth=2)
regr_2=DecisionTreeRegressor(max_depth=5)
regr_3=DecisionTreeRegressor(max_depth=7)
regr_1.fit(X, distance_list)
regr_2.fit(X, distance_list)
regr_3.fit(X, distance_list)

X_test = np.arange(0.0, steps, 0.01)[:, np.newaxis]
y_1 = regr_1.predict(X_test)
y_2 = regr_2.predict(X_test)
y_3=regr_3.predict(X_test)

# Plot the results
plt.figure()
plt.scatter(X, distance_list, s=2.5, c="black", label="data")
plt.plot(X_test, y_1, color="red",
         label="max_depth=2", linewidth=2)
plt.plot(X_test, y_2, color="green", label="max_depth=5", linewidth=2)
plt.plot(X_test, y_3, color="m", label="max_depth=7", linewidth=2)

plt.xlabel("Data")
plt.ylabel("Darget")
plt.title("Decision Tree Regression")
plt.legend()
plt.show()

\end{minted}

% !split
\subsection*{Artificial neurons}

The field of artificial neural networks has a long history of
development, and is closely connected with the advancement of computer
science and computers in general. A model of artificial neurons was
first developed by McCulloch and Pitts in 1943 to study signal
processing in the brain and has later been refined by others. The
general idea is to mimic neural networks in the human brain, which is
composed of billions of neurons that communicate with each other by
sending electrical signals.  Each neuron accumulates its incoming
signals, which must exceed an activation threshold to yield an
output. If the threshold is not overcome, the neuron remains inactive,
i.e.~has zero output.

This behaviour has inspired a simple mathematical model for an artificial neuron.

\[
 y = f\left(\sum_{i=1}^n w_ix_i\right) = f(u)
\]
Here, the output $y$ of the neuron is the value of its activation function, which have as input
a weighted sum of signals $x_i, \dots ,x_n$ received by $n$ other neurons.
% --- end paragraph admon ---

% !split
\subsection*{A simple perceptron model}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{figures/perceptron.png}}

% !split
\subsection*{Neural network types}

% --- begin paragraph admon ---
\paragraph{}
An artificial neural network (NN), is a computational model that consists of layers of connected neurons, or \emph{nodes}. 
It is supposed to mimic a biological nervous system by letting each neuron interact with other neurons
by sending signals in the form of mathematical functions between layers. 
A wide variety of different NNs have
been developed, but most of them consist of an input layer, an output layer and eventual layers in-between, called
\emph{hidden layers}. All layers can contain an arbitrary number of nodes, and each connection between two nodes
is associated with a weight variable.

% inline figure
\centerline{\includegraphics[width=0.6\linewidth]{figures/dnn.png}}

% !split
\subsection*{The system: two electrons in a harmonic oscillator trap in two dimensions}

@@

% --- begin paragraph admon ---
\paragraph{Basic steps.}
Choose a trial wave function
$\psi_T(\bm{R})$.
\[
   P(\bm{R},\bm{\alpha})= \frac{\left|\psi_T(\bm{R},\bm{\alpha})\right|^2}{\int \left|\psi_T(\bm{R},\bm{\alpha})\right|^2d\bm{R}}.
\]
This is our model, or likelihood/probability distribution function  (PDF). It depends on some variational parameters $\bm{\alpha}$.
The approximation to the expectation value of the Hamiltonian is now 
\[
   \langle E[\bm{\alpha}] \rangle = 
   \frac{\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})H(\bm{R})\Psi_T(\bm{R},\bm{\alpha})}
        {\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})\Psi_T(\bm{R},\bm{\alpha})}.
\]
% --- end paragraph admon ---

@@

% !split
\subsection*{The VMC code}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
# Importing various packages
from math import exp, sqrt
from random import random, seed
import numpy as np
import matplotlib.pyplot as plt
from mpl_toolkits.mplot3d import Axes3D
from matplotlib import cm
from matplotlib.ticker import LinearLocator, FormatStrFormatter
import sys

#Trial wave function for quantum dots in two dims
def WaveFunction(r,alpha,beta):
    r1 = r[0,0]**2 + r[0,1]**2
    r2 = r[1,0]**2 + r[1,1]**2
    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
    deno = r12/(1+beta*r12)
    return exp(-0.5*alpha*(r1+r2)+deno)

#Local energy  for quantum dots in two dims, using analytical local energy
def LocalEnergy(r,alpha,beta):

    r1 = (r[0,0]**2 + r[0,1]**2)
    r2 = (r[1,0]**2 + r[1,1]**2)
    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
    deno = 1.0/(1+beta*r12)
    deno2 = deno*deno
    return 0.5*(1-alpha*alpha)*(r1 + r2) +2.0*alpha + 1.0/r12+deno2*(alpha*r12-deno2+2*beta*deno-1.0/r12)

# The Monte Carlo sampling with the Metropolis algo
def MonteCarloSampling():

    NumberMCcycles= 100000
    StepSize = 1.0
    # positions
    PositionOld = np.zeros((NumberParticles,Dimension), np.double)
    PositionNew = np.zeros((NumberParticles,Dimension), np.double)
    # seed for rng generator
    seed()
    # start variational parameter
    alpha = 0.9
    for ia in range(MaxVariations):
        alpha += .025
        AlphaValues[ia] = alpha
        beta = 0.2 
        for jb in range(MaxVariations):
            beta += .01
            BetaValues[jb] = beta
            energy = energy2 = 0.0
            DeltaE = 0.0
            #Initial position
            for i in range(NumberParticles):
                for j in range(Dimension):
                    PositionOld[i,j] = StepSize * (random() - .5)
            wfold = WaveFunction(PositionOld,alpha,beta)

            #Loop over MC MCcycles
            for MCcycle in range(NumberMCcycles):
                #Trial position
                for i in range(NumberParticles):
                    for j in range(Dimension):
                        PositionNew[i,j] = PositionOld[i,j] + StepSize * (random() - .5)
                wfnew = WaveFunction(PositionNew,alpha,beta)

                #Metropolis test to see whether we accept the move
                if random() < wfnew**2 / wfold**2:
                   PositionOld = PositionNew.copy()
                   wfold = wfnew
                   DeltaE = LocalEnergy(PositionOld,alpha,beta)
                energy += DeltaE
                energy2 += DeltaE**2

            #We calculate mean, variance and error ...
            energy /= NumberMCcycles
            energy2 /= NumberMCcycles
            variance = energy2 - energy**2
            error = sqrt(variance/NumberMCcycles)
            Energies[ia,jb] = energy    
    return Energies, AlphaValues, BetaValues

#Here starts the main program with variable declarations
NumberParticles = 2
Dimension = 2
MaxVariations = 10
Energies = np.zeros((MaxVariations,MaxVariations))
AlphaValues = np.zeros(MaxVariations)
BetaValues = np.zeros(MaxVariations)
(Energies, AlphaValues, BetaValues) = MonteCarloSampling()

# Prepare for plots
fig = plt.figure()
ax = fig.gca(projection='3d')
# Plot the surface.
X, Y = np.meshgrid(AlphaValues, BetaValues)
surf = ax.plot_surface(X, Y, Energies,cmap=cm.coolwarm,linewidth=0, antialiased=False)
# Customize the z axis.
zmin = np.matrix(Energies).min()
zmax = np.matrix(Energies).max()
ax.set_zlim(zmin, zmax)
ax.set_xlabel(r'$\alpha$')
ax.set_ylabel(r'$\beta$')
ax.set_zlabel(r'$\langle E \rangle$')
ax.zaxis.set_major_locator(LinearLocator(10))
ax.zaxis.set_major_formatter(FormatStrFormatter('%.02f'))
# Add a color bar which maps values to colors.
fig.colorbar(surf, shrink=0.5, aspect=5)
plt.show()

\end{minted}

% !split
\subsection*{Technical aspect, improvements and how to define the cost function}

\textbf{The above procedure is not the smartest one}. Looping over all variational parameters becomes expensive.
Also, we don't use importance sampling and optimizations of the standard deviation (blocking, bootstrap, jackknife). 
Such codes are included in the above Github address.

We can also be smarter and use minimization methods to find the \textbf{optimal} variational parameters with fewer Monte Carlo cycles and then 
fire up our heavy artillery.

One way to achieve this is to minimize the energy as function of the variational parameters.
% --- end paragraph admon ---

% !split
\subsection*{Energy derivatives}

% --- begin paragraph admon ---
\paragraph{}
To find the derivatives of the local energy expectation value as function of the variational parameters, we can use the chain rule and the hermiticity of the Hamiltonian.

Let us define 
\[
\bar{E}_{\alpha_i}=\frac{d\langle  E_L\rangle}{d\alpha_i}.
\]
as the derivative of the energy with respect to the variational parameter $\alpha_i$
We define also the derivative of the trial function (skipping the subindex $T$) as 
\[
\bar{\Psi}_{i}=\frac{d\Psi}{d\alpha_i}.
\]
% --- end paragraph admon ---

% !split
\subsection*{Derivatives of the local energy}

@@

How should we define the cost function?
% --- end paragraph admon ---

% !split
\subsection*{Meet the variance and its derivatives}

% --- begin paragraph admon ---
\paragraph{Why the variance?}
Suppose the trial function (our model) is the exact wave function. The action of the hamiltionan
on the wave function
\[
   H\Psi = \mathrm{constant}\times \Psi,
\]
The integral which defines various 
expectation values involving moments of the hamiltonian becomes then
\[
   \langle E^n \rangle =   \langle H^n \rangle =
   \frac{\int d\bm{R}\Psi^{\ast}(\bm{R})H^n(\bm{R})\Psi(\bm{R})}
        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=
\mathrm{constant}\times\frac{\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}
        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=\mathrm{constant}.
\]
\textbf{This gives an important information: If I want the variance, the exact wave function leads to zero variance!}
The variance is defined as 
\[
\sigma_E = \langle E^2\rangle - \langle E\rangle^2.
\]
Variation is then performed by minimizing both the energy and the variance.
% --- end paragraph admon ---

% !split
\subsection*{The variance defines the cost function}

% --- begin paragraph admon ---
\paragraph{}
We can then take the derivatives of 
\[
\sigma_E = \langle E^2\rangle - \langle E\rangle^2,
\]
with respect to the variational parameters. The derivatives of the variance can then be used to defined the
so-called Hessian matrix, which in turn allows us to use minimization methods like Newton's method or 
standard gradient methods.

This leads to however a more complicated expression, with obvious errors when evaluating integrals by Monte Carlo integration. Less used, see however \href{{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.94.150201}}{Filippi and Umrigar}. The expression becomes complicated
\[
\bar{E}_{ij} = 2\left[ \langle (\frac{\bar{\Psi}_{ij}}{\Psi}+\frac{\bar{\Psi}_{j}}{\Psi}\frac{\bar{\Psi}_{i}}{\Psi})(E_L-\langle E\rangle)\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\bar{E}_j-\langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\bar{E}_i\right] +\langle \frac{\bar{\Psi}_{i}}{\Psi}E_L{_j}\rangle +\langle \frac{\bar{\Psi}_{j}}{\Psi}E_L{_i}\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L{_j}\rangle \langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\langle E_L{_i}\rangle.  
\]

Evaluating the cost function means having to evaluate the above second derivative of the energy.
% --- end paragraph admon ---

>>>>

@@

\section{Introduction and motivation}

Traditional many-body methods like full configuration interaction
theory (FCI), coupled-cluster (CC) theory, many-body perturbation
theory (MBPT), in-medium similarity renormalization group (IMSRG),
various Monte Carlo methods and many other many-body approaches, have
been rather successful in describing properties of nuclear
systems. However, essentially all of these methods face what is
normally called the curse of dimensionality. For wave function based
method like FCI or CC theories where the original continuous problem
is discretized in terms selected basis functions and/or many-body
excitations, the dimensionality of systems under study grows almost
exponentially when larger basis sets and/or number of particles are
included. Nuclear systems close to the limits of stability pose in
particular a tough problem to many-body practitioners in terms of an
increase of the number of relevant degrees of freedom. To describe say
weakly bound nuclei within the framework of a wave function based
approach, requires often a single-particle basis which includes bound,
weakly bound and unbound states, increasing thereby considerably the
number of many-body basis states. This renders often a standard
shell-model calculation infeasible. Within the above mentioned
many-body methods there are however several interesting theoretical
approachs which attempt at circumventing the dimensionality
curse. Smarter basis sets is one of these approaches, as well as the
resummation of specific correlations and the recently proposed
stochastic sampling of many-body states in both FCI and CC
calculations.

Recently, several authors of pointed to  possibilities within the
broad fields of Machine Learning and Quantum Computing as approaches
that hold great promise in tackling the ever increasing
dimensionalities. There are several groups worldwide which now focus
on Machine Learning and/or Quantum Computing applied to many-body
problems.

@@

Here we try to implement these algorithms to many-body systems that mimick a nuclear physics like problem, namely electrons confined to move in harmonic oscillator like potentials.

Machine learning (ML) is an extremely rich field, in spite of its young age. The
increases we have seen during the last three decades in computational
capabilities have been followed by developments of methods and
techniques for analyzing and handling large date sets, relying heavily
on statistics, computer science and mathematics.  The field is rather
new and developing rapidly.

Almost every problem in ML and data science starts with the same ingredients:
\begin{itemize}
\item The dataset $\mathbf{x}$ (could be some observable quantity of the system we are studying)

@@

\noindent
We seek to minimize the function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ by finding the parameter values which minimize $\mathcal{C}$. This leads to  various minimization algorithms.

\section{What is Machine Learning}
Machine learning is the science of giving computers the ability to
learn without being explicitly programmed.  The idea is that there
exist generic algorithms which can be used to find patterns in a broad
class of data sets without having to write code specifically for each
problem. The algorithm will build its own logic based on the data.

@@

You will notice however that many of the basic ideas discussed do come from Physics!

The approaches to machine learning are many, but are often split into two main categories. 
In \emph{supervised learning} we know the answer to a problem,
and let the computer deduce the logic behind it. On the other hand, \emph{unsupervised learning}
is a method for finding patterns and relationship in data sets without any prior knowledge of the system.
Some authours also operate with a third category, namely \emph{reinforcement learning}. This is a paradigm 
of learning inspired by behavioural psychology, where learning is achieved by trial-and-error, 
solely from rewards and punishment.

@@

\begin{itemize}
\item An excellent reference, \href{{https://arxiv.org/abs/1803.08823}}{Mehta \emph{et al.}, arXiv:1803.08823 and Physics Reports in press (2018)}

@@

@@

\subsection*{The system: two electrons in a harmonic oscillator trap in two dimensions}

@@

\subsection*{Quantum Monte Carlo Motivation}

\subsection*{Quantum Monte Carlo Motivation}

@@

@@

How should we define the cost function?
